Analysis of Monte Carlo accelerated iterative methods\\ for sparse linear systems

### 2017-03-06 doc/nla/mcsa.tex
Added NLA article sources.

%
% This draft dated January 29, 2016  [MB]
%

\documentclass[final,leqno,onefignum,onetabnum]{siamltex1213}
\usepackage{amssymb,amsmath,bm}
\usepackage{color}
\usepackage[lined,boxruled]{algorithm2e}
\usepackage{comment}
\usepackage{calrsfs}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage{rotating}
\usepackage{multirow}
\DeclareMathAlphabet{\pcal}{OMS}{zplm}{m}{n}

 \newtheorem{thm}{Theorem}
% \newtheorem{lemma}{Lemma}
 \newtheorem{defn}{Definition}
 \newtheorem{prop}{Proposition}
\newtheorem{rem}{Remark}

\hoffset 0.6in

%opening
\title{Analysis of Monte Carlo accelerated iterative methods for
sparse linear systems \tnoteref{ornl-cr}}
\author{Michele Benzi\thanks{Department of Mathematics and Computer
Science, Emory University, Atlanta, Georgia 30322, USA
(benzi@mathcs.emory.edu). The work of this author was supported
by Department of Energy (Office of Science) grant ERKJ247.}
\and
Thomas M.~Evans\thanks{Oak Ridge National Laboratory, 1 Bethel Valley Rd.,
Oak Ridge, TN 37831, USA (evanstm@ornl.gov).}
\and Steven P. Hamilton\thanks{Oak Ridge National Laboratory,
1 Bethel Valley Rd., Oak Ridge, TN 37831, USA (hamiltonsp@ornl.gov).}
\and
\linebreak Massimiliano Lupo  Pasini\thanks{Department of Mathematics and Computer
Science, Emory University, Atlanta, Georgia 30322, USA (mlupopa@emory.edu).}
\and
Stuart R.~Slattery\thanks{Oak Ridge National Laboratory,
1 Bethel Valley Rd., Oak Ridge, TN 37831, USA (slatterysr@ornl.gov).}
}
\tnotetext[ornl-cr]{Notice: This manuscript
  has been authored by UT-Battelle, LLC, under contract DE-AC05-00OR22725
  with the U.S. Department of Energy. The United States Government retains
  and the publisher, by accepting the article for publication, acknowledges
  that the United States Government retains a non-exclusive, paid-up,
  irrevocable, world-wide license to publish or reproduce the published form
  of this manuscript, or allow others to do so, for United States Government
  purposes.}
\date{}

\begin{document}
\bibliographystyle{siam}

\pagestyle{myheadings}
\markboth{{\sc M.~Benzi, T.~M.~Evans, S.~P.~Hamilton, M.~Lupo Pasini, and S.~Slattery}}
{Monte Carlo solvers for sparse linear systems}

@@

\begin{keywords}
sparse linear systems, iterative methods, Richardson iteration, preconditioning,
Monte Carlo methods, sparse approximate inverses, resilience
\end{keywords}

\begin{AMS}
65C05, 65F50, 65F08, 65F10
\end{AMS}

@@

Numerical evidence presented in \cite{EMSH2014} suggests
that MCSA can be competitive, for certain classes of problems,
with established deterministic solvers such as preconditioned
conjugate gradients and GMRES. So far, however, no theoretical analysis
of the convergence properties of these solvers has been carried out. In
particular, it is not clear a priori whether the method, applied to a
particular linear system, will converge.
Indeed, the convergence of the underlying preconditioned Richardson
iteration is not sufficient, in general, to guarantee the convergence
of the MCSA-accelerated iteration. In other words, it is quite possible
that the stochastic ``acceleration" part of the algorithm may actually
cause the hybrid method to diverge or stagnate.

@@

By restricting the attention to a single component of $\mathbf{x}$ we
obtain
\begin{equation}
x_i=f_i + \sum_{\ell=1}^\infty \sum_{k_1=1}^n\sum_{k_2=1}^n\cdots
\sum_{k_{\ell}=1}^n
H_{i,k_1}H_{k_1,k_2}\cdots H_{k_{\ell-1}, k_{\ell}}f_{k_{\ell}}.
\label{forward}
\end{equation}
The last equation can be reinterpreted as the realization of an estimator
defined on a random walk.  Let us start considering a random walk whose
state space $S$ is labeled by the set of indices of the forcing term
$\mathbf{f}$:
\[
S=\{1,2,\ldots, n\} \subset \mathbb{N}.
\]
Each $i$th step of the random walk has a random variable
$k_i$ associated with it. The realization of $k_i$ represents the index of the
component of $\mathbf{f}$
which is visited in the current step of the random walk.
The transition probabilities and the selection of
the initial state of the random walk can be accomplished according to
different modalities, leading to two distinct methods:
the \textit{forward} and \textit{adjoint} methods.
These methods are described next.

@@

The general goal is to evaluate a functional as
\[
J(\mathbf{x})=\left \langle\mathbf{h},\mathbf{x}\right\rangle=\sum_{i=1}^n h_i
x_i,
\]
where $\mathbf{h}\in \mathbb{R}^n$ is the Riesz representative in
$\mathbb{R}^n$ of the functional $J$.
Such a representative can be used to build the initial probability
$\tilde{p}:
S\rightarrow [0,1]$ of the random walk as
\[
\tilde{p}(k_0=i)=\tilde{p}_{k_0}=\frac{\lvert h_i\rvert}{\sum_{i=1}^n
\lvert h_i\rvert}.
\]
It is important to highlight that the role of vector $\mathbf{h}$ is
confined to the construction of the initial probability, and that $\mathbf{h}$
is not
used afterwards in the stochastic process.
A possible choice for the transition probability $P$ can be
\[
p(k_{\ell}=j \;\lvert\;k_{\ell-1}=i )=P_{ij}=\frac{\lvert
H_{ij}\rvert}{\sum_{k=1}^n
\lvert H_{ik}\rvert}.
\]
where
\[
p(\cdot,i):S\rightarrow [0,1], \qquad \forall i\in S
\]
and $k_{\ell}\in S$
represents the state reached at a generic $\ell$th step of the random walk.
A related sequence of random variables $w_{ij}$ can be defined
as
\[
w_{ij}=\frac{H_{ij}}{P_{ij}}.
\]
The probability distribution of the random variables $w_{ij}$ is represented
by the transition matrix that governs the stochastic process. The $w_{ij}$
quantities just introduced can be used to build one more sequence
of random variables.
At first we introduce quantities $W_\ell$
\[
W_{0}=\frac{h_{k_0}}{\tilde{p}_{k_0}}, \quad W_{\ell}=W_{\ell-1}
w_{k_{\ell-1},k_{\ell}}.
\]
By introducing $\nu$ as a generic permutation that represents the realization
of a random walk, we define
\[
X(\nu)=\sum_{\ell=0}^\infty W_{\ell} f_{k_{\ell}}
\]
as the random variable associated with a specific permutation $\nu$.
We can thus
define the estimator $\theta$ as
\[
\theta=E[X]=\sum_{\nu}P_{\nu}X(\nu),
\]
where $\nu$ ranges over all possible realizations.
$P_{\nu}$ is the probability associated with a specific permutation of the
random walk.
It can be proved (see \cite{Hal1962} and \cite{Hal1994}) that
\[
E[W_{\ell}
f_{k_{\ell}}]=\left\langle\mathbf{h},H^{\ell}\mathbf{f}\right\rangle, \quad
\ell=0,1,2,\ldots
\]
and
\[
\theta=E\bigg[\sum_{\ell=0}^\infty W_{\ell}
f_{k_{\ell}}\bigg]=\left\langle\mathbf{h},\mathbf{x}\right\rangle.
\]

@@

The forward method presented above, however, has the limitation of employing an
entire set of permutations to estimate just a single entry of
the solution at a time. Hence, in order to estimate the entire solution
vector for Eq.~\eqref{linsys}, we have to employ a separate set of
permutations for each entry in the solution vector. This limitation
can be circumvented by the adjoint method which we describe below.

@@

The forward method adds a contribution to the component of the solution
vector where the random walk began, based on the value of the source vector
in the state in which the walk currently resides.  The adjoint method,
on the other hand, adds a contribution to the component of the solution
vector where the random walk currently resides based on the value of the
source vector in the state in which the walk began.
The Kronecker delta at the end of the series \eqref{adj_mean} represents a
filter, indicating
that only a subset of permutations contribute to the $j$th component
of the solution vector.

@@

The main advantage of this method, compared to the forward one, consists in the
fact that a single set of permutations is used to estimate the entire solution
vector.  Unless only a small portion of the problem is of interest, this
property often leads to the adjoint method being favored over the forward
method.
 In other terms, the adjoint method should be preferred when approximating
the solution globallly over the entire computational domain, while the forward
method is especially useful when approximating the solution locally.

@@

Hence, the expected value estimator averages
the deterministic contribution of the iteration matrix over all the potential
states $j$ that might be reached from the current state $\ell$. The variance
in this case becomes:
\begin{equation}
Var\bigg [\sum_{\ell=0}^\infty W_{\ell}
H_{k_{\ell},i}^T\bigg]=E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
{H_{k_{\ell},i}^T}^T\bigg ] - \bigg
(E\bigg[\sum_{\ell=0}^\infty
W_{\ell}
H_{k_{\ell},i}^T\bigg]\bigg )^2, \qquad i=1,\ldots,n
\label{adj_var1}.
\end{equation}

@@

\begin{algorithm}[H]
 %\SetAlgoLined
 %\BlankLine
 \LinesNumbered
 \KwData{$H$, $\mathbf{b}$, $\mathbf{x}_0$}
 \KwResult{$\mathbf{x}_{num}$}
 $l=0$\;
 \While{not reached convergence}{
  $\mathbf{r}^{l}=\mathbf{b}-A\mathbf{x}^{l}$\;
  $\delta \mathbf{x}^{l+1}\approx(I-H)^{-1}\mathbf{r}^{l}$; \qquad
\text \% {Computed via Standard MC}
  $\mathbf{x}^{l+1}=\mathbf{x}^{l}+\delta \mathbf{x}^{l+1}$\;
  $l=l+1$\;
 }
 $\mathbf{x}_{num}=\mathbf{x}^{l+1}$\;
 \caption{Sequential Monte Carlo}
\end{algorithm}

The Monte Carlo linear solver method is used to compute the update $\delta
\mathbf{x}^{l}$.  This algorithm is equivalent to a Richardson iteration
accelerated by a correction obtained by approximately solving the error-residual
equation
\begin{equation}
(I - H)\delta \mathbf{x}^{l+1} = \mathbf{r}^{l}.
\label{eq:corr}
\end{equation}
If this equation were to be solved exactly, the corresponding approximation
$\mathbf{x}^{l+1}=\mathbf{x}^{l}+\delta \mathbf{x}^{l+1}$ would be the exact
solution to the linear system. This is of course impractical, since solving
(\ref{eq:corr}) is equivalent to solving the original linear system
(\ref{linsys}). Instead, the correction is obtained by
solving (\ref{eq:corr}) only approximately, using a Monte Carlo method.
Because Monte Carlo is only applied within a single iteration, the central
limit theorem is only applicable within that iteration rather than to
the overall convergence behavior of the algorithm. This allows a trade-off
between the amount of time and effort spent on the inner (stochastic) and
outer (deterministic) iterations, which can take account the competing goal
of reliability and rapid convergence.

@@

\begin{algorithm}[H]
\LinesNumbered
 \KwData{$H$, $\mathbf{b}$, $\mathbf{x}_0$}
 \KwResult{$\mathbf{x}_{num}$}
 $l=0$\;
 \While{not reached convergence}{
  $\mathbf{x}^{l+\frac{1}{2}}=H\mathbf{x}^l+\mathbf{b}$\;
  $\mathbf{r}^{l+\frac{1}{2}}=\mathbf{b}-A\mathbf{x}^{l+\frac{1}{2}}$\;
 $\delta\mathbf{x}^{l+\frac{1}{2}}\approx(I-H)^{-1}\mathbf{r}^{l+\frac{1}{2}}$;
\qquad
\% \text{Computed via Standard MC}
  $\mathbf{x}^{l+1}=\mathbf{x}^{l+\frac{1}{2}}+\delta
\mathbf{x}^{l+\frac{1}{2}}$\;
 $l=l+1$\;
 }
 $\mathbf{x}_{num}=\mathbf{x}^{l+1}$\;
 \caption{Monte Carlo Synthetic Acceleration}
\end{algorithm}

@@

On the one hand, sufficient conditions for
convergence of Monte Carlo linear solvers are very restrictive; see, e.g.,
\cite{Srin2010} and \cite{MASC2013}. On the other hand, the necessary and
sufficient
condition in \cite{MASC2013} requires knowledge of $\rho(\hat{H})$, which
is not readily available. Note that explicit computation of $\rho(\hat{H})$
is quite expensive, comparable to the cost of solving the original linear system.
While ensuring that $\rho(H)<1$ (by means with appropriate preconditioning)
is in many cases possible, guaranteeing that $\rho(\hat{H})<1$ is generally
much more problematic.

@@

Note that this means that there exists a diagonal matrix  $\Delta$
such that $A\Delta$ is SDD.
A proper subclass of the class of GDD matrices is represented
by the nonsingular $M$-matrices.
Recall that $A$ is a nonsingular $M$-matrix if it is of the form
$A=r I - B$ where $B$ is nonnegative and $r>\rho(B)$.
It can be shown (see, e.g., \cite{BP1979}) that  a matrix
$A\in\mathbb{R}^{n\times n}$ is a nonsingular $M$-matrix if and only
if there exists a positive diagonal matrix $\Delta$ such that
$A\Delta$ is SDD by rows. Clearly, every
nonsingular $M$-matrix is GDD.

@@

Nevertheless, if $A$ is an $M$-matrix there exist efficient methods
to determine a diagonal scaling of $A$ so that the resulting matrix
is SDD by rows. Note that the scaled matrix is still an $M$-matrix,
therefore applying left Jacobi preconditioning to this matrix will
guarantee that both $\rho(H)<1$ and $\rho(\hat{H})<1$.

In \cite{Li2002}, the author presents a procedure to determine
whether a given matrix $A\in\mathbb{C}^{n\times n}$
is GDD (in which case the diagonal scaling that
makes $A$ GDD is produced), or not.
The algorithm can be described as follows.

@@

One possible preconditioning approach involves either diagonal or block
diagonal preconditioning (with blocks of small or moderate size). Diagonal
preconditioning does not alter the sparsity of the original coefficient matrix,
while block diagonal preconditioning will incur a moderate amount of
fill-in in the preconditioned matrix if the blocks are not too large.
From the discussions in Section~\ref{sec:sdd}
and \ref{sec:bdd}, selecting a diagonal or block diagonal preconditioner
guarantees convergence of the Monte Carlo schemes
for matrices that are strictly diagonally dominant
or block diagonally dominant, respectively. In addition, $M$-matrices
that are not strictly or block diagonally dominant can also be dealt with by
first rescaling $A$ so that it becomes strictly diagonally dominant, as
discussed in Section~\ref{sec:sdd}.

@@

>>>>

\documentclass[times]{nlaauth}

\usepackage{moreverb}

\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
\usepackage{amssymb,amsmath,bm}
\usepackage{color}
\usepackage[ruled]{algorithm2e}
\usepackage{comment}
\usepackage{calrsfs}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage{rotating}
\usepackage{multirow}
\DeclareMathAlphabet{\pcal}{OMS}{zplm}{m}{n}

 \newtheorem{thm}{Theorem}
 \newtheorem{lemma}{Lemma}
 \newtheorem{defn}{Definition}
 \newtheorem{prop}{Proposition}
\newtheorem{rem}{Remark}

%\hoffset 0.6in

%opening
%\title{Analysis of Monte Carlo accelerated iterative methods for
%sparse linear systems}
%\author{Michele Benzi\thanks{Department of Mathematics and Computer
%Science, Emory University, Atlanta, Georgia 30322, USA
%(benzi@mathcs.emory.edu). The work of this author was supported
%by Department of Energy (Office of Science) grant ERKJ247.}
%\and
%Thomas M.~Evans\thanks{Oak Ridge National Laboratory, 1 Bethel Valley Rd.,
%Oak Ridge, TN 37831, USA (evanstm@ornl.gov).}
%\and Steven P. Hamilton\thanks{Oak Ridge National Laboratory,
%1 Bethel Valley Rd., Oak Ridge, TN 37831, USA (hamiltonsp@ornl.gov).}
%\and
%\linebreak Massimiliano Lupo  Pasini\thanks{Department of Mathematics and Computer
%Science, Emory University, Atlanta, Georgia 30322, USA (mlupopa@emory.edu).}
%\and
%Stuart R.~Slattery\thanks{Oak Ridge National Laboratory,
%1 Bethel Valley Rd., Oak Ridge, TN 37831, USA (slatterysr@ornl.gov).}
%}

\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\def\volumeyear{2017}
\def\DOI{nla.2008}

\begin{document}

\runningheads{M.~Benzi et al.}{Monte Carlo solvers for sparse linear systems}

\title{Analysis of Monte Carlo accelerated iterative methods\\ for
sparse linear systems}

\author{Michele Benzi\affil{1}, Thomas M.~Evans\affil{2}, Steven P. Hamilton\affil{3},
\linebreak Massimiliano Lupo  Pasini\affil{4}, and Stuart R.~Slattery\affil{5}
}

\address{\affilnum{1}Department of Mathematics and Computer
Science, Emory University, Atlanta, Georgia 30322, USA (benzi@mathcs.emory.edu).\break
\affilnum{2}Oak Ridge National Laboratory, 1 Bethel Valley Rd.,
Oak Ridge, TN 37831, USA (evanstm@ornl.gov).\break
\affilnum{3}Oak Ridge National Laboratory, 1 Bethel Valley Rd.,
Oak Ridge, TN 37831, USA (hamiltonsp@ornl.gov).\break
\affilnum{4}Department of Mathematics and Computer
Science, Emory University, Atlanta, Georgia 30322, USA (mlupopa@emory.edu).\break
\affilnum{5}Oak Ridge National Laboratory, 1 Bethel Valley Rd.,
Oak Ridge, TN 37831, USA (slatterysr@ornl.gov).}

\cgsn{\affilnum{1}This manuscript has been authored by UT-Battelle, LLC under
  Contract No. DE-AC05-00OR22725 with the U.S. Department of Energy. The
  United States Government retains and the publisher, by accepting the article
  for publication, acknowledges that the United States Government retains a
  non-exclusive, paid-up, irrevocable, world-wide license to publish or
  reproduce the published form of this manuscript, or allow others to do so,
  for United States Government purposes. The Department of Energy will provide
  public access to these results of federally sponsored research in accordance
  with the DOE Public Access Plan
  (http://energy.gov/downloads/doe-public-access-plan).}{Dept. of Energy
  (Office of Science) ERKJ247}

\pagestyle{myheadings}
%\markboth{{\sc M.~Benzi, T.~M.~Evans, S.~P.~Hamilton, M.~Lupo Pasini, and S.~R.~Slattery}}
%{Monte Carlo solvers for sparse linear systems}

@@

\keywords{sparse linear systems; iterative methods; Richardson iteration; preconditioning;
Monte Carlo methods; sparse approximate inverses; resilience}

%\begin{AMS}
%65C05, 65F50, 65F08, 65F10
%\end{AMS}

@@

Numerical evidence presented in \cite{EMSH2014} suggests
that MCSA can be competitive, for certain classes of problems,
with established deterministic solvers such as preconditioned
conjugate gradients and Generalized Minimum Residual (GMRES). So far, however, no theoretical analysis
of the convergence properties of these solvers has been carried out. In
particular, it is not clear a priori whether the method, applied to a
particular linear system, will converge.
Indeed, the convergence of the underlying preconditioned Richardson
iteration is not sufficient, in general, to guarantee the convergence
of the MCSA-accelerated iteration. In other words, it is quite possible
that the stochastic ``acceleration" part of the algorithm may actually
cause the hybrid method to diverge or stagnate.

@@

By restricting the attention to a single component of $\mathbf{x}$ we
obtain
\begin{equation}
x_i=f_i + \sum_{\ell=1}^\infty \sum_{k_1=1}^n\sum_{k_2=1}^n\cdots
\sum_{k_{\ell}=1}^n
H_{i,k_1}H_{k_1,k_2}\cdots H_{k_{\ell-1}, k_{\ell}}f_{k_{\ell}}.
\label{forward}
\end{equation}
The last equation can be reinterpreted as the realization of an estimator
defined on a random walk.  Let us start considering a random walk whose
state space $S$ is labeled by the set of indices of the forcing term
$\mathbf{f}$:
\[
S=\{1,2,\ldots, n\} \subset \mathbb{N}.
\]
Each $i$th step of the random walk has a random variable
$k_i$ associated with it. The realization of $k_i$ represents the index of the
component of $\mathbf{f}$
which is visited in the current step of the random walk.
 The construction of random walks is accomplished considering the directed graph associated with the matrix $H$.
The nodes of this graph are labeled 1 through $n$, and there is a directed edge from node $i$ to node $j$
if and only if $H_{i,j}\ne 0$. Starting from a given node, the random walk consists of a sequence of nodes obtained by
jumping from one node to the next along directed edges, choosing the next node at random according to a transition
probability distribution matrix constructed from $H$ or from $H^T$, see below. Note that it may happen that a row of $H$
(or of $H^T$) is all zero; this happens
when there are no out-going (respectively, in-coming) edges from (respectively, to)
the corresponding node. In this case, that particular random walk is terminated and another
node is selected as the starting point for the next walk.
The transition probabilities and the selection of
the initial state of the random walk can be accomplished according to
different modalities, leading to two distinct methods:
the \textit{forward} and \textit{adjoint} methods.
These methods are described next.

@@

 Given a functional $J$, the goal is to compute its action on a vector $\mathbf{x}$ by
 constructing a statistical estimator whose expected value equals $J(\mathbf{x})$. Each statistical sampling
is represented by a random walk and it contributes to build an estimate of the expected value.
Towards this goal, it is necessary to introduce an initial probability distribution
and a transition matrix so that random walks are well defined.
Recalling Riesz's representation theorem one can write
\[
J(\mathbf{x})=\left \langle\mathbf{h},\mathbf{x}\right\rangle=\sum_{i=1}^n h_i
x_i,
\]
where $\mathbf{h}\in \mathbb{R}^n$ is the Riesz representative in
$\mathbb{R}^n$ of the functional $J$.
Such a representative can be used to build the initial probability
$\tilde{p}:
S\rightarrow [0,1]$ of the random walk as
\[
\tilde{p}(k_0=i)=\tilde{p}_{k_0}=\frac{\lvert h_i\rvert}{\sum_{i=1}^n
\lvert h_i\rvert}.
\]
It is important to highlight that the role of vector $\mathbf{h}$ is
confined to the construction of the initial probability, and that $\mathbf{h}$
is not
used afterwards in the stochastic process.
A possible choice for the transition probability matrix $P$ can be
\[
p(k_{\ell}=j \;\lvert\;k_{\ell-1}=i )=P_{ij}=\frac{\lvert
H_{ij}\rvert}{\sum_{k=1}^n
\lvert H_{ik}\rvert}.
\]
where
\[
p(\cdot,i):S\rightarrow [0,1], \qquad \forall i\in S
\]
and $k_{\ell}\in S$
represents the state reached at a generic $\ell$th step of the random walk.
A related sequence of random variables $w_{ij}$ can be defined
as
\[
w_{ij}=\frac{H_{ij}}{P_{ij}}.
\]
The probability distribution of the random variables $w_{ij}$ is represented
by the transition matrix that governs the stochastic process. The $w_{ij}$
quantities just introduced can be used to build one more sequence
of random variables.
At first we introduce quantities $W_\ell$
\[
W_{0}=\frac{h_{k_0}}{\tilde{p}_{k_0}}, \quad W_{\ell}=W_{\ell-1}
w_{k_{\ell-1},k_{\ell}}.
\]
In probability theory, a random walk is itself envisioned as a random variable that can assume multiple
values consisting of different realizations. Indeed, given a starting point, there are in general many choices
(one for each nonzero in the corresponding row of $P$) to select the second state and from there on recursively.
The actual feasibility of a path and the frequency with which it is selected depend on the initial probability and on
the transition matrix.
By introducing $\nu$ as a realization
of a random walk, we define
\[
X(\nu)=\sum_{\ell=0}^\infty W_{\ell} f_{k_{\ell}}
\]
as the random variable associated with a specific realization $\nu$.
We can thus
define the estimator $\theta$ as
\[
\theta=E[X]=\sum_{\nu}P_{\nu}X(\nu),
\]
where $\nu$ ranges over all possible realizations.
$P_{\nu}$ is the probability associated with a specific realization of the
random walk.
It can be proved (see \cite{Hal1962} and \cite{Hal1994}) that
\[
E[W_{\ell}
f_{k_{\ell}}]=\left\langle\mathbf{h},H^{\ell}\mathbf{f}\right\rangle, \quad
\ell=0,1,2,\ldots
\]
and
\[
\theta=E\bigg[\sum_{\ell=0}^\infty W_{\ell}
f_{k_{\ell}}\bigg]=\left\langle\mathbf{h},\mathbf{x}\right\rangle.
\]

@@

The forward method presented above, however, has the limitation of employing an
entire set of realizations to estimate just a single entry of
the solution at a time. Hence, in order to estimate the entire solution
vector for Eq.~\eqref{linsys}, we have to employ a separate set of
realizations for each entry in the solution vector. This limitation
can be circumvented by the adjoint method which we describe below.

@@

The forward method adds a contribution to the component of the solution
vector where the random walk began, based on the value of the source vector
in the state in which the walk currently resides.  The adjoint method,
on the other hand, adds a contribution to the component of the solution
vector where the random walk currently resides based on the value of the
source vector in the state in which the walk began.
The Kronecker delta at the end of the series \eqref{adj_mean} represents a
filter, indicating
that only a subset of realizations contribute to the $j$th component
of the solution vector.

@@

The main advantage of this method, compared to the forward one, consists in the
fact that a single set of realizations is used to estimate the entire solution
vector.  Unless only a small portion of the problem is of interest, this
property often leads to the adjoint method being favored over the forward
method.
 In other terms, the adjoint method should be preferred when approximating
the solution globally over the entire computational domain, while the forward
method is especially useful when approximating the solution locally.

@@

Hence, the expected value estimator averages
the deterministic contribution of the iteration matrix over all the potential
states $j$ that might be reached from the current state $\ell$. The variance
in this case becomes:
\begin{equation}
Var\bigg [\sum_{\ell=0}^\infty W_{\ell}
H_{k_{\ell},i}^T\bigg]=E\bigg[\sum_{\ell=0}^\infty W_{\ell}^2
{H_{k_{\ell},i}^T}\bigg ] - \bigg
(E\bigg[\sum_{\ell=0}^\infty
W_{\ell}
H_{k_{\ell},i}^T\bigg]\bigg )^2, \qquad i=1,\ldots,n
\label{adj_var1}.
\end{equation}

@@

\begin{algorithm}[H]
 %\SetAlgoLined
 %\BlankLine
 \LinesNumbered
 \KwData{$H$, $\mathbf{b}$, $\mathbf{x}_0$}
 \KwResult{$\mathbf{x}_{num}$}
 $l=0$\;
 \While{not reached convergence}{
  $\mathbf{r}^{l}=\mathbf{b}-A\mathbf{x}^{l}$\;
  $\delta \mathbf{x}^{l+1}\approx(I-H)^{-1}\mathbf{r}^{l}$; \qquad
\text \% {Computed via Standard MC}\;
  $\mathbf{x}^{l+1}=\mathbf{x}^{l}+\delta \mathbf{x}^{l+1}$\;
  $l=l+1$\;
 }
 $\mathbf{x}_{num}=\mathbf{x}^{l+1}$\;
 \caption{Sequential Monte Carlo}
\end{algorithm}

The Monte Carlo linear solver method is used to compute the update $\delta
\mathbf{x}^{l}$.  This algorithm is equivalent to a Richardson iteration
accelerated by a correction obtained by approximately solving the error-residual
equation
\begin{equation}
(I - H)\delta \mathbf{x}^{l+1} = \mathbf{r}^{l}.
\label{eq:corr}
\end{equation}
If this equation were to be solved exactly, the corresponding approximation
$\mathbf{x}^{l+1}=\mathbf{x}^{l}+\delta \mathbf{x}^{l+1}$ would be the exact
solution to the linear system. This is of course impractical, since solving
(\ref{eq:corr}) is equivalent to solving the original linear system
(\ref{linsys}). Instead, the correction is obtained by
solving (\ref{eq:corr}) only approximately, using a Monte Carlo method.
Because Monte Carlo is only applied within a single iteration, the central
limit theorem is only applicable within that iteration rather than to
the overall convergence behavior of the algorithm. This allows a trade-off
between the amount of time and effort spent on the inner (stochastic) and
outer (deterministic) iterations, which can take into account the competing goals
of reliability and rapid convergence.

@@

\begin{algorithm}[H]
\LinesNumbered
 \KwData{$H$, $\mathbf{b}$, $\mathbf{x}_0$}
 \KwResult{$\mathbf{x}_{num}$}
 $l=0$\;
 \While{not reached convergence}{
  $\mathbf{x}^{l+\frac{1}{2}}=H\mathbf{x}^l+\mathbf{b}$\;
  $\mathbf{r}^{l+\frac{1}{2}}=\mathbf{b}-A\mathbf{x}^{l+\frac{1}{2}}$\;
 $\delta\mathbf{x}^{l+\frac{1}{2}}\approx(I-H)^{-1}\mathbf{r}^{l+\frac{1}{2}}$;
\qquad
\% \text{Computed via Standard MC}\;
  $\mathbf{x}^{l+1}=\mathbf{x}^{l+\frac{1}{2}}+\delta
\mathbf{x}^{l+\frac{1}{2}}$\;
 $l=l+1$\;
 }
 $\mathbf{x}_{num}=\mathbf{x}^{l+1}$\;
 \caption{Monte Carlo Synthetic Acceleration}
\end{algorithm}

@@

On the one hand, sufficient conditions for
convergence of Monte Carlo linear solvers are very restrictive; see, e.g.,
\cite{Srin2010} and \cite{MASC2013}. On the other hand, the necessary and
sufficient
condition in \cite{MASC2013} requires knowledge of $\rho(\hat{H})$, which
is not readily available. Note that explicit computation of $\rho(\hat{H})$
is quite expensive, comparable to the cost of solving the original linear system.
While ensuring that $\rho(H)<1$ (by means of appropriate preconditioning)
is in many cases possible, guaranteeing that $\rho(\hat{H})<1$ is generally
much more problematic.

@@

A proper subclass of the class of GDD matrices is represented
by the nonsingular $M$-matrices.
Recall that $A$ is a nonsingular $M$-matrix if it is of the form
$A=r I - B$ where $B$ is nonnegative and $r>\rho(B)$.
It can be shown (see, e.g., \cite{BP1979}) that  a matrix
$A\in\mathbb{R}^{n\times n}$ is a nonsingular $M$-matrix if and only
if there exists a positive diagonal matrix $\Delta$ such that
$A\Delta$ is SDD by rows. Clearly, every
nonsingular $M$-matrix is GDD.

@@

Nevertheless, if $A$ is an $M$-matrix, there exist efficient methods
to determine a diagonal scaling of $A$ so that the resulting matrix
is SDD by rows. Note that the scaled matrix is still an $M$-matrix,
therefore applying left Jacobi preconditioning to this matrix will
guarantee that both $\rho(H)<1$ and $\rho(\hat{H})<1$.

In \cite{Li2002}, the author presents a procedure to determine
whether a given matrix $A\in\mathbb{C}^{n\times n}$
is GDD (in which case the diagonal scaling that
makes $A$ SDD is produced), or not.
The algorithm can be described as follows.

@@

One possible preconditioning approach involves either diagonal or block
diagonal preconditioning (with blocks of small or moderate size). Diagonal
preconditioning does not alter the sparsity of the original coefficient matrix,
whereas block diagonal preconditioning will incur a moderate amount of
fill-in in the preconditioned matrix if the blocks are not too large.
From the discussions in Section~\ref{sec:sdd}
and \ref{sec:bdd}, selecting a diagonal or block diagonal preconditioner
guarantees convergence of the Monte Carlo schemes
for matrices that are strictly diagonally dominant
or block diagonally dominant, respectively. In addition, $M$-matrices
that are not strictly or block diagonally dominant can also be dealt with by
first rescaling $A$ so that it becomes strictly diagonally dominant, as
discussed in Section~\ref{sec:sdd}.

@@

\subsection{Considerations about computational complexity} \label{sec:complex}

Providing an analysis of the computational complexity for the aforementioned algorithms is not entirely
straightforward because of their stochastic nature. Indeed, different statistical samplings can produce estimates with
different uncertainty levels, requiring a proper tuning of the number of samplings computed to reach a prescribed accuracy.
Moreover, we already mentioned that the asymptotic analysis of MC convergence assumes random walks with infinitely many steps
and $N\rightarrow \infty$, where $N$ is the number of random walks. However, in practice each history must be truncated to a
finite number of steps and the number of statistical samplings must be finite as well. The actual value of $N$ and the length
of the histories affect the accuracy of the statistical estimation, thus influencing the number of iterations in a hybrid algorithm,
since the uncertainty propagates to subsequent iterations.
Therefore, here  we can only provide a tentative analysis of the complexity of the forward and adjoint Monte Carlo
methods, assuming a specific history length and a fixed number $N$ of statistical samplings.

Recalling formula \eqref{dir_mean} for the entry-wise estimate for the solution with the forward method, the cost of
reconstructing the entire solution vector is
\begin{equation}
\text{Forward:}\quad \mathcal{O}(N\cdot k_{\ell}\cdot n),
\label{for_cost}
\end{equation}
where $N$ is the number of histories, $k_{\ell}$ is the length of each random walk, and $n$ is the number of unknowns.
As concerns the adjoint method, formula \eqref{adj_mean} leads to
\begin{equation}
\text{Adjoint:}\quad \mathcal{O}(N\cdot k_{\ell}).
\label{adj_cost}
\end{equation}
The operation count for the hybrid schemes can thus be obtained by combining the cost of the Richardson scheme with the
complexity of standard MC techniques.

Regarding the Sequential Monte Carlo algorithm, the standard MC scheme is combined with the computation of the residual
at each iteration. The cost of the residual computation is essentially that of a sparse matrix-vector product, which is
$\mathcal{O}(n)$ for a sparse system. Therefore, the complexity of a single Sequential MC iteration is
\begin{equation}
\text{Sequential MC with forward MC update:}\quad \mathcal{O}((n + 1) \cdot N\cdot k_{\ell})
\label{seqfor_cost}
\end{equation}
using the forward MC as an inner solver, and
\begin{equation}
\text{Sequential MC with adjoint MC update:}\quad \mathcal{O}(n + N\cdot k_{\ell})
\label{seqadj_cost}
\end{equation}
using the adjoint MC as an inner solver.

A single iteration of MCSA requires computing the residual, applying a matrix-vector product using $H$, and applying an MC update.
The complexity of an MCSA iteration using the forward MC is therefore
\begin{equation}
\text{MCSA with forward MC update:}\quad \mathcal{O}((2n + 1) \cdot N\cdot k_{\ell}),
\label{mcsafor_cost}
\end{equation}
whereas using the adjoint MC we find
\begin{equation}
\text{MCSA with adjoint MC update:}\quad \mathcal{O}(2n + N\cdot k_{\ell}).
\label{mcsaadj_cost}
\end{equation}
Note that these estimates require $nnz(H)\approx nnz(A)$, in the sense that both $H$ and $A$ contain $\mathcal{O}(n)$ nonzeros.
The actual values attained by $N$ and $k_{\ell}$ depend on the thresholds employed to truncate a single history and to determine
the number of random walks to use. In general, the higher $N$ and $k_{\ell}$, the lower the number of outer iterations to achieve
a prescribed accuracy.
Finally, the total cost will depend also on the number of iterations, which is difficult to predict in practice.
